Compilation through interpretation

### 2023-07-17 Papers/FEDCIS-2023/cm1.tex
corrections from FEDCIS review process

>>>>

\begin{abstract}

	As static metaprogramming is becoming more relevant, compilers must adapt to accommodate them. This requires exposing more information about the code, from the compiler to the programmer as well as more powerful compile-time function execution capabilities. The Interpreter component of a compiler therefore becomes more important.

	In this paper a novel approach to compiler architecture that places the Interpreter as the central component of the compiler is proposed.
	Translation of user code into the executable form is done by an Interpreter, written in the target language. The data structures of Interpreter are accessible also to the programmer. Such solution significantly improves flexibility and extensibility of the compiler and enables execution of any code at the compile-time.

	Compile Time Function Execution (CTFE) First pattern was designed for low-level, non-garbage-collected, reflection-enabled language C-=-1. This is a new programming language, which allows the programmer to execute any code at the compile time, as well as to analyze and modify the program structure. Grace to the extensibility of the designed compiler, programs written in C-=-1 can also generate marshalling bindings for other languages and support a variety of programming paradigms.

	The significant flexibility and extensibility of CTFEF caused severe problems in compiler construction. The compiler appeared very complex, its parts, especially Interpreter, were very difficult to debug. Compiler operates on inflexible data structures, accessible also for a programmer. They are therefore a part of the compiled languages standard library and backwards compatibility must be maintained.

\end{abstract}

### 2023-07-17 Papers/FEDCIS-2023/sections/conclusions.tex
corrections from FEDCIS review process

On the other hand, using CTFEF approach has some drawbacks. Implementing a compiler is more difficult.
Since a significant part of the compiler is interpreted, the initial implementation requires working with the target language.
The lack of available tools, such as integrated development environments, debuggers and libraries makes this part of the process significantly more difficult.
On the other hand also means the work on the compiler in the target language may start at the very beginning of compiler bootstrapping process\cite{puntambekar:compiler_design, novillo2007gcc}.

>>>>

On the other hand, using CTFEF approach has some drawbacks. Implementing a compiler is more difficult.
Since a significant part of the compiler is interpreted, the initial implementation requires working with the target language.
The lack of available tools, such as integrated development environments, debuggers and libraries, for this new language, makes this part of the process significantly more difficult.
On the other hand also means the work on the compiler in the target language may start at the very beginning of compiler bootstrapping process\cite{puntambekar:compiler_design, novillo2007gcc}.

### 2023-07-17 Papers/FEDCIS-2023/sections/design.tex
fixed article
corrections from FEDCIS review process

C-=-1 is a simple language, built with minimum set of features needed to demonstrate the usefulness of the proposed metaprogramming features.
They are discussed further in chapter  \ref{design:attributes_and_metaprogramming}.
The primary motivation for those mechanisms was to provide to a programmer the ability to create domain specific static analysis and code generation tools, without creating a separate program.

@@

Listing \ref{lst:noDiscardCm1} contains an example of a C-=-1 attribute providing static analysis: \lstinline{noDiscard}.
It works in the same manner as the attribute of the same name present in C++17 \cite{ISO:cpp17}: if applied to a function, the result of that invoking procedure must be used.
To declare an attribute in C-=-1, the \lstinline{att} keyword is used.
After that, attribute targets should be listed in angled brackets, as in line 0 of Listing \ref{lst:noDiscardCm1}.
Valid targets for attributes include a number of language elements, such as types, functions, variables and fields. Attribute from Listing \ref{lst:noDiscardCm1} declares two member functions: \lstinline{attach} in line 4 and \lstinline{onCall} in line 6.

@@

CTFEF approach was created during implementation of the first compiler for the C-=-1 language\cite{grabski2022compilation}.
CTFEF compiler has four major components:
\begin{enumerate}
	\item Frontend
	\item Interpreter
	\item Compiler Interface
	\item Backend
\end{enumerate}

Figure \ref{CTFE-first-compiler-structure} contains a diagram with an overview of how these parts interact with each other, during the compilation process.
Frontend, described in chapter \ref{Frontend}, parses the code in the compiled language and constructs its intermediate representation, using Interpreter's data structures.
It is used to analyze both user code and the Compiler Interface.
After the intermediate representation is constructed, it is passed to the Interpreter, which is described in chapter \ref{Interpreter}.
Compiler Interface intermediate representation is then executed, using the user program as data.
This step converts the semantic model of the program into the Backends intermediate language.
This process is further explained in chapter \ref{compiler-interface}.
Finally, the Backend generates the executable file.

\begin{figure}
	\includegraphics[height=14cm]{pictures/compiler-structure.png}
	\caption{CTFEF compiler structure}
	\label{CTFE-first-compiler-structure}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=7cm]{pictures/circular-function-reference.jpg}
	\caption{Example of circular reference in meta code}
	\label{circular-function-reference}
\end{figure}

An important issue for CTFEF compiler is decision what code can be executed at compile-time.
In some languages it may be possible to introduce circular references between the functions that modify the codebase.
Figure \ref{circular-function-reference} contains a diagram of such scenario.
\lstinline{A}, \lstinline{B} and \lstinline{C} are functions.
If function \lstinline{B} is modified by function \lstinline{C} and \lstinline{C} invokes \lstinline{B}, the behavior of function \lstinline{A} is unpredictable.
This problem will only be magnified by larger program sizes.

@@

Compiler Interface translates the program's intermediate representation into the Backend's assembly language.
This component is interpreted during compilation.

>>>>

C-=-1 is a simple language, built with minimum set of features needed to demonstrate the usefulness of the proposed metaprogramming features.
They are discussed further in section  \ref{design:attributes_and_metaprogramming}.
The primary motivation for those mechanisms was to provide to a programmer the ability to create domain specific static analysis and code generation tools, without creating a separate program.

@@

Listing \ref{lst:noDiscardCm1} contains an example of a C-=-1 attribute providing static analysis: \lstinline{noDiscard}.
It works in the same manner as the attribute of the same name present in C++17 \cite{ISO:cpp17}: the result of invoking the annotated function must be used.
To declare an attribute in C-=-1, the \lstinline{att} keyword is used.
After that, attribute targets should be listed in angled brackets, as in line 0 of Listing \ref{lst:noDiscardCm1}.
Valid targets for attributes include a number of language elements, such as types, functions, variables and fields. Attribute from Listing \ref{lst:noDiscardCm1} declares two member functions: \lstinline{attach} in line 4 and \lstinline{onCall} in line 6.
Listing \ref{lst:noDiscardUsageCm1} contains an example of using the \lstinline{noDiscard} attribute.
All uses of the \lstinline|noDiscardFunction|, except for the one on line 3 are valid.
Using the function in a statement as opposed to an expression causes a compile-time error.

@@

CTFEF approach was created during implementation of the first compiler for the C-=-1 language\cite{grabski2022compilation}.
CTFEF compiler has four major components:
\begin{enumerate}
	\item Frontend.
	\item Interpreter.
	\item Compiler Interface.
	\item Backend.
\end{enumerate}

Figure \ref{CTFE-first-compiler-structure} contains a diagram with an overview of how these parts interact with each other, during the compilation process.
Frontend, described in section \ref{Frontend}, parses the code in the compiled language and constructs its intermediate representation, using Interpreter's data structures.
The structure and form of the intermediate representation is not specified by the CTFEF approach, as long as it contains all semantically relevant information from the source program.
It is used to analyze both user code and the Compiler Interface.
After the intermediate representation is constructed, it is passed to the Interpreter, which is described in section \ref{Interpreter}.
Compiler Interface intermediate representation is then executed, using the user program as data.
This step converts the semantic model of the program into the Backends intermediate language.
This process is further explained in section \ref{compiler-interface}.
Finally, the Backend generates the executable file.

\begin{figure}
	\centering
	\includegraphics[height=13cm]{pictures/compiler-structure.png}
	\caption{CTFEF compiler structure}
	\label{CTFE-first-compiler-structure}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=7cm]{pictures/circular-function-reference.jpg}
	\caption{Example of circular reference in meta code}
	\label{circular-function-reference}
\end{figure}

An important issue for CTFEF compiler is decision what code can be executed at compile-time.
In some languages it may be possible to introduce circular references between the functions that modify the codebase.
Figure \ref{circular-function-reference} contains a diagram of such scenario.
\lstinline{A}, \lstinline{B} and \lstinline{C} are functions.
If function \lstinline{B} is modified by function \lstinline{C} and \lstinline{B} invokes \lstinline{C}, the behavior of function \lstinline{A} is unpredictable.
This problem will only be magnified by larger program sizes.

@@

Compiler Interface translates the program's intermediate representation into the Backend's assembly language.
This component is interpreted during compilation and may be provided by the user.
In case of C-=-1, compiler interface could be supplied to the compiler, the same way that the code being compiled is passed, as a collection of source files.
Figure \ref{compiler-design} reflects this decision, treating the Compiler Interface as an input into the compiler, same as user code.

### 2023-07-17 Papers/FEDCIS-2023/sections/implementation.tex
fixed article
corrections from FEDCIS review process

The first CTFEF was created for a new language: C-=-1, using generic parser generator and Backend.
The most important aspect of implementing a CTFEF compiler is the design of the data structures, described in chapter \ref{data_structures}, that will be used by the Interpreter.

In order to exploit CTFEF approach in design of a compiler the language should contain a set of data structures to describe user code i.e. a Semantic Model.
It will allow the programmer to interact and manipulate the structure of the program at compile-time.
The Semantic model designed and implemented for C-=-1 is described in chapter \ref{semantic_model}.

The final part of a CTFEF compiler is the Backend Interface.
It is a program, written in the target language, and executed at compile-time that translates the semantic model into the Backends' assembly language.
Backend Interface implemented for C-=-1 is relatively small and is described in chapter \ref{implementation/Backend-interface}.

\begin{figure*}
	\includegraphics[width=15cm]{pictures/Interpreter_data_structures_uml.png}
	\caption{Class diagram of C-=-1 Interpreter data structures}
	\label{fig:Interpreter_data_structures}
\end{figure*}

>>>>

The first CTFEF was created for a new language: C-=-1, using generic parser generator and Backend.
The most important aspect of implementing a CTFEF compiler is the design of the data structures, described in section \ref{data_structures}, that will be used by the Interpreter.

In order to exploit CTFEF approach in design of a compiler the language should contain a set of data structures to describe user code i.e. a Semantic Model.
It will allow the programmer to interact and manipulate the structure of the program at compile-time.
The Semantic model designed and implemented for C-=-1 is described in section \ref{semantic_model}.

The final part of a CTFEF compiler is the Backend Interface.
It is a program, written in the target language, and executed at compile-time that translates the semantic model into the Backends' assembly language.
Backend Interface implemented for C-=-1 is relatively small and is described in section \ref{implementation/Backend-interface}.

\begin{figure*}
	\centering
	\includegraphics[width=13cm]{pictures/Interpreter_data_structures_uml.png}
	\caption{Class diagram of C-=-1 Interpreter data structures}
	\label{fig:Interpreter_data_structures}
\end{figure*}

### 2023-07-17 Papers/FEDCIS-2023/sections/introduction.tex
corrections from FEDCIS review process

Compile-time function execution (CTFE) is an aspect of a programming language, that allows the programmer to execute code at compile-time.
It has been gaining popularity with programming languages without virtual machines such as Rust \cite{rust} or C++ \cite{ISO:cpp98}.
The goals and capabilities of CTFE depend on the language and are discussed in details in chapter \ref{related-work}.

In this paper a new approach to compiler construction that places Compile Time Function Execution capabilities as the first priority is proposed.
This architecture is called CTFEF: Compile Time Function Execution First. CTFEF builds the compiler around the Interpreter component.
The role of the compiler is to construct the semantic model of the program being compiled and pass it as data to the interpreted Compiler-Interface module.
Compiler-Interface then transforms the model into the assembly language to be compiled into executable.
This approach was created during implementation of the compiler for C-=-1, a new language that prioritizes static metaprogramming.
The design of that language is described in chapter \ref{language-design}.
Related work is briefly shown in chapter \ref{related-work}.
In chapter \ref{compiler-design} the structure of a CTFEF compiler and interactions among its components are presented.

Using CTFEF approach causes many implementation difficulties. Data structures used within the compiler are tightly coupled with the compiled language.
These problems and how they were solved in C-=-1 compiler are described in chapter \ref{implementation}.

>>>>

Compile-time function execution (CTFE) is an aspect of a programming language, that allows the programmer to execute code at compile-time.
It has been gaining popularity with programming languages without virtual machines such as Rust \cite{rust} or C++ \cite{ISO:cpp98}.
The goals and capabilities of CTFE depend on the language and are discussed in details in section \ref{related-work}.

In this paper a new approach to compiler construction that places Compile Time Function Execution capabilities as the first priority is proposed.
This architecture is called CTFEF: Compile Time Function Execution First. CTFEF builds the compiler around the Interpreter component.
The goal of this approach is to shorten the initial stage of compiler bootstrapping \cite{puntambekar:compiler_design, novillo2007gcc}, better support languages built around static metaprogramming and make the compiler more extensible.
The role of the compiler is to construct the semantic model of the program being compiled and pass it as data to the interpreted Compiler-Interface module.
Compiler-Interface then transforms the model into the assembly language to be compiled into executable.
This approach was created during implementation of the compiler for C-=-1, a new language that prioritizes static metaprogramming.
Using CTFEF approach causes many implementation difficulties. Data structures used within the compiler are tightly coupled with the compiled language.
These problems and how they were solved in C-=-1 compiler are described in section \ref{implementation}.

The design of that language is described in section \ref{language-design}.
Related work is briefly shown in section \ref{related-work}.
In section \ref{compiler-design} the structure of a CTFEF compiler and interactions among its components are presented.
